Concurrent Programming in Harmony

### 2021-02-23 book/paper.tex
Latest

\autoref{fig:uplock} showed a faulty attempt at solving mutual
exclusion using a lock.  The problem with the implementation of the
lock is that checking the lock and setting it if it is available is
not \emph{atomic}.  Thus multiple threads contending for the lock
can all ``grab the lock'' at the same time.  While Peterson's
algorithm gets around the problem, it is not efficient, especially
if generalized to multiple threads.
Worse, most modern processors make accessing shared memory locations
non-atomic to allow for efficient caching, out-of-order execution,
and other optimizations.
And even compilers may re-order memory accesses when optimizing code.
%
Peterson's algorithm relies on a sequential consistent memory model,
and hence the \textbf{sequential} statement: without it Harmony will
complain about data races.
We saw in \autoref{ex:reorder} that a seemingly harmless reordering
of instructions can break the code.
The memory models of C and Java are not sequentially consistent, and
they do not provide a \textbf{sequential} statement either, so one
cannot implement Peterson's algorithm in those languages very easily.

@@

\index{data race}
When there is a bug in a program because some code omitted obtaining
a lock before accessing the data structure, that is known as a
\emph{data race}.
In Harmony, a data race has a specific meaning: a state in which multiple
threads are trying to access the same variable, and at least one of those
accesses updates the variable.
However, there are two exceptions.
First, using the \textbf{sequential} statement, you can specify that
concurrent access to the specified variables is intended.
Second, if the accesses are within an \textbf{atomic} statement, then they
are not considered part of a data race.
However, for the same reason that you should not use \textbf{atomic}
statements in your ordinary synchronization code, you should not use
\textbf{sequential} statements either.  Neither is supported by regular
programming languages.  But both can be very useful for testing.

>>>>

\autoref{fig:uplock} showed a faulty attempt at solving mutual
exclusion using a lock.  The problem with the implementation of the
lock is that checking the lock and setting it if it is available is
not \emph{atomic}.  Thus multiple threads contending for the lock
can all ``grab the lock'' at the same time.  While Peterson's
algorithm gets around the problem, it is not efficient, especially
if generalized to multiple threads.
Worse, most modern processors make accessing shared memory locations
non-atomic to allow for efficient caching, out-of-order execution,
and other optimizations.
And even compilers may re-order memory accesses when optimizing code.
%
Peterson's algorithm relies on a sequential consistent memory model,
and hence the \textbf{sequential} statement: without it Harmony will
complain about data races.
We saw in \autoref{ex:reorder} that a seemingly harmless reordering
of instructions can break the code.
The memory models of C and Java are not sequentially consistent, and
they do not provide a \textbf{sequential} statement either, so one
cannot implement Peterson's algorithm in those languages directly.

@@

\index{data race}
When there is a bug in a program because some code omitted obtaining
a lock before accessing a shared data structure, that is known as a
\emph{data race}.
More precisely, a data race happens when there is a state in which multiple
threads are trying to access the same variable, and at least one of those
accesses updates the variable.
In many environments, including C and Java programs, the behavior of
concurrent load and store operations have tricky or even undefined
semantics.  One should therefore avoid data races, which is why Harmony
reports them even though Harmony has sequentially consistent memory.

Harmony does not report data races in two cases.
First, using the \textbf{sequential} statement, you can specify that
concurrent access to the specified variables is intended.
Second, if the accesses are within an \textbf{atomic} statement, then they
are not considered part of a data race.
For the same reason that you should not use \textbf{atomic}
statements in your ordinary synchronization code, you should not use
\textbf{sequential} statements either.  Neither is supported by regular
programming languages.  They exist in Harmony for the purpose of
defining interlock instructions and for testing.
The statements should only appear in modules that define synchronization
primitives or in test programs.
